Efficiently stable presentations from error-correcting codes

### 2023-10-01 efficient_stability.tex
some responses to intro comments

\paragraph{Motivation.}
A linear error-correcting code $\code$ is a $k$-dimensional subspace of the vector space $\F^n$ over a finite field $\F$ that has certain combinatorial properties. The foremost of these is the \emph{minimal distance} $d$, which is defined as the smallest Hamming weight (number of nonzero coordinates) $|c|$ of a nonzero vector  $c\in\code$. A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that \hnote{was: uniquely identify elements} that identifies $\code$ within $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho\, d(x,\code)$, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$. The terminology ``locally testable'' comes from an interpretation of $\frac{1}{m}|hx|$ as the probability of rejection of a natural ``tester'' for $\code$, i.e.\ an algorithm that on input $x$ checks a randomly chosen row $h_i$ of $h$ and accepts if and only if $h_i\cdot x = \sum_j h_{ij} x_j =0$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. In general one would like to design families of codes of increasing length $n$, such that $d$ is a linear function of $n$ and $\rho$ is bounded below by a constant independent of $n$. \hnote{Should this be promoted to the main text? Feels a bit odd that it's a footnote.}}  Thus a code is called locally testable if words that are far from the code have a high probability of being rejected according to this tester. This notion plays a central role in applications of codes to complexity theory~\cite{babai1991non,PCP_thm}, and continues to be actively studied, see e.g.~\cite{LTC_DELLM,LTC_Panteleev_Kalachev} for a recent breakthrough on the topic.

@@

\paragraph{Group stability.}
This observation, which at first may seem quite anodyne or even forced, opens a universe of questions \hnote{how about: ``This observation immediately raises many questions.''?}. The problem of relating approximate representations of a group to exact representations of it is termed \emph{stability} in group theory, and has a long history. There are of course many flavors of the problem, depending on how one defines closeness (should it be the operator norm or the Hilbert-Schmidt norm? Should closeness hold for every relation, or every pair of group elements, or is it sufficient that it holds on average? Etc.) We will review some relevant results in this area below. For now, we mention Voiculescu's famous counter-example~\cite{voiculescu1983asymptotically} about approximately commuting unitaries, which was motivated by a question of Halmost on pairs of approximately commuting Hermitian operators~\cite{halmos1976some}. Using the terminology of stability, Voiculescu showed that the presentation $\Z^2 = \langle x,y:[x,y]=e\rangle$ is \emph{not} stable with respect to the operator norm. However, much more recently Glebsky~\cite{glebsky2010almost} showed that the same presentation \emph{is} stable with respect to the normalized Hilbert-Schmidt norm.

This example and many others show that the notion of stability is, in general, highly sensitive to the notion of closeness considered. Returning to our main concern, so far we have argued that 
stability of approximate \emph{$1$-dimensional} representations of the presentation $G(h)$ is connected to local testability of the code $\ker h$. In the case of $1$-dimensional representations of course the choice of norm does not matter; however, the choice of measuring the error on average over relations, as opposed to e.g.\ taking the maximum, is one in which we depart from most of the literature. We will motivate this choice further below; but before we can continue we must pause to introduce the key definitions that our work builds on. For the purposes of the introduction we focus the discussion on finite-dimensional representations; in the main paper we handle the general case of representations in a tracial von Neumann algebra. For $X\in \C^{d\times d}$ let $\|X\|_f^2 = \frac{1}{d}\Tr(X^* X)$ denote the normalized Hilbert-Schmidt norm, \hnote{added:} let $\mU(\C^d)$ denote the unitary operators on $\C^d$, and for a set $S$ let $\mF(S)$ denote the free group generated by the elements of $S$.

\begin{definition}[Almost homomorphism]\label{def:approx-hom-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group and $\mu_R$ a distribution on $R$. An $(\eps,\mu_R)$-almost homomorphism of $G$ is a homomorphism $\phi:\mF(S)\to\mU(\C^d)$ for some $d\geq 1$ such that
\[ \Es{r\sim \mu_R} \big\|  \phi(r) - \Id \big \|_f^2 \,\leq\, \eps\;.\]
Here, $\phi(r)$ denotes the product of $\phi(s)$ for $s$ appearing in $r$, in the specified order. 
{\color{red} YOU ALREADY SAY THAT YOU ARE GIVEN A HOMOMOTPHISM FROM THE FREE GROUP. SO $\phi(r)$ IS WELL DEFINED... NO NEED FOR THE REMARK "Here ...." AT THE END}.
\end{definition}

As already mentioned this definition makes two importance choices: firstly, to measure closeness in the Hilbert-Schmidt norm, and secondly, to measure it on average over the choice of a relation. Next we give our definition for a finitely presented group to be stable.

@@

\begin{remark}
\hnote{promoted this from footnote to remark} It is natural, and will always be the case for us, to take $\mu_S$ as the ``marginal'' of $\mu_R$, i.e.\ the distribution on $S$ obtained by sampling $r\sim\mu_R$ and returning a uniformly random $s$ that appears in $r$, when $r$ is considered as a multi-set.
\end{remark}

@@

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? One can verify that with $\mu_R$ and $\mu_S$ the uniform distribution on $R$ and $S$ respectively, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for some $\delta=O(\eps)$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$ \hnote{what's $\rho$? it should be related to $\delta$ somehow..}. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. However, the answer is not trivial. In particular, and perhaps surprisingly, the group $G(h)$ is not necessarily Abelian, see Remark~\ref{rk:non-abelian}. \hnote{this seems like an indirect way to motivate why it's nontrivial... why does non-Abelian-ness mean it's difficult? is there a better way of justifying the non-triviality of this question?}

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question \hnote{I'm not so convinced we've properly justified its difficulty... }, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the most part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$ (which are the only groups we consider in this paper) the multiplication table presentation \hnote{is this standard terminology?}, which has $|G|$ generators and $|G|^2$ relations, is $\delta(\eps)=C\eps$-stable \hnote{hmm... technically the notion of stability in GH is not the same as the one defined above, right?} for some constant $C$ that is independent of the group. This holds even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability alluded to above.

\hnote{edited this paragraph a bit; the mention of $\Z_2^k$ seemed to pop out of nowhere, but I don't think this discussion about efficiency relies on it being $\Z_2^k$.} However the results of~\cite{gowers2017inverse,de2019operator} do not imply the same quantitative stability bounds for $G(h)$ with respect to its defining presentation~\eqref{eq:gh-intro} (even if we assume that $G(h)$ is finite \hnote{could it always be finite?}). The reason to prefer the presentation~\eqref{eq:gh-intro} as opposed to the multiplication table presentation of $G(h)$ is that~\eqref{eq:gh-intro} is much more succinct: if $n,k$ are linearly related then it has $O(k)$ generators and relations, as opposed to $2^k$ and $2^{2k}$ respectively. Such a gain is essential when one recalls the interpretation of the presentation $G(h)$ as a ``tester'' for $G$ --- the size of the presentation is then directly related to the amount of \emph{randomness} required by the tester (to sample a random relation); and in computer science applications randomness is seen as an essential resource (we discuss this more below, in the context of quantum computing).

@@

\begin{theorem*}[Main, informal]
For every integer $k\geq 1$ there is a presentation $\Z_2^k = \langle S_k:R_k\rangle$ such that $|S_k|,|R_k| = 2^{\poly\log(k)}$ and furthermore this presentation is $\delta(\eps)=\poly(\log k,\eps)$-stable. \hnote{are we assuming some default setting for $\mu_S,\mu_R,d$, etc?}
\end{theorem*}

@@

\begin{remark}\label{rk:non-abelian}
There exists matrices $h$ such that $G(h)$ is not $\Z_2^k$, and in fact is not Abelian. For an example, see~\cite[Example 2.16]{paddock2022arkhipov}. For that example, $n=12$, $k=7$, and the parity check matrix $h$ can be described explicitly as follows: the $7$ rows of $h$ are indexed by the vertices of the complete bipartite graph $K_{3,4}$, the $12$ columns are indexed by the $12$ edges of $K_{3,4}$, and the entry $(i,j)$ of $h$ is $1$ if and only if the edge $j$ is incident on vertex $i$. As shown in~\cite{paddock2022arkhipov}, $G(h)$ is not Abelian. 
\end{remark}

>>>>

\paragraph{Motivation.}
A linear error-correcting code $\code$ is a $k$-dimensional subspace of the vector space $\F^n$ over a finite field $\F$ that has certain combinatorial properties. The foremost of these is the \emph{minimal distance} $d$, which is defined as the smallest Hamming weight (number of nonzero coordinates) $|c|$ of a nonzero vector  $c\in\code$. In general one would like to design families of codes of increasing length $n$, such that both $k$ and $d$ are bounded below by a positive linear function of $n$. Such codes are referred to as ``good'' codes.

A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that specify $\code$ as a subspace of $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho\, d(x,\code)$, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$.  Ideally one would like to design families of good codes such that in addition $\rho$ is bounded below by a constant independent of $n$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. So a family of codes can be ``locally testable'' without being ``good.'' \hnote{Should this be promoted to the main text? Feels a bit odd that it's a footnote.}\tnote{Yes, I moved some of it to the text in two different places. It was doing a couple things at once. The remaining part is less important.} } This is a challenging task, and the construction of families of good locally testable codes was only achieved very recently~\cite{LTC_DELLM}.

The terminology ``locally testable'' comes from an interpretation of $\frac{1}{m}|hx|$ as the probability of rejection of a natural ``tester'' for $\code$, i.e.\ an algorithm that on input $x$ checks a randomly chosen row $h_i$ of $h$ and accepts if and only if $h_i\cdot x = \sum_j h_{ij} x_j =0$.  Thus a code is called locally testable if words that are far from the code have a high probability of being rejected according to this tester. This notion plays a central role in applications of codes to complexity theory~\cite{babai1991non,PCP_thm}, and continues to be actively studied, see e.g.~\cite{LTC_DELLM,LTC_Panteleev_Kalachev} for a recent breakthrough on the topic.

@@

\paragraph{Group stability.}
This observation immediately raises many questions. The problem of relating approximate representations of a group to exact representations of it is termed \emph{stability} in group theory, and has a long history. There are of course many flavors of the problem, depending on how one defines closeness (should it be the operator norm or the Hilbert-Schmidt norm? Should closeness hold for every relation, or every pair of group elements, or is it sufficient that it holds on average? Etc.) We will review some relevant results in this area below. For now, we mention Voiculescu's famous counter-example~\cite{voiculescu1983asymptotically} about approximately commuting unitaries, which was motivated by a question of Halmost on pairs of approximately commuting Hermitian operators~\cite{halmos1976some}. Using the terminology of stability, Voiculescu showed that the presentation $\Z^2 = \langle x,y:[x,y]=e\rangle$ is \emph{not} stable with respect to the operator norm. However, much more recently Glebsky~\cite{glebsky2010almost} showed that the same presentation \emph{is} stable with respect to the normalized Hilbert-Schmidt norm.

This example and many others show that the notion of stability is, in general, highly sensitive to the notion of closeness considered. Returning to our main concern, so far we have argued that 
stability of approximate \emph{$1$-dimensional} representations of the presentation $G(h)$ is connected to local testability of the code $\ker h$. In the case of $1$-dimensional representations of course the choice of norm does not matter; however, the choice of measuring the error on average over relations, as opposed to e.g.\ taking the maximum, is one in which we depart from most of the literature. We will motivate this choice further below; but before we can continue we must pause to introduce the key definitions that our work builds on. For the purposes of the introduction we focus the discussion on finite-dimensional representations; in the main paper we handle the general case of representations in a tracial von Neumann algebra. Let $\mU(\C^d)$ denote the unitary operators on $\C^d$, and for a set $S$ let $\mF(S)$ denote the free group generated by the elements of $S$. For $X\in \C^{d\times d}$ let $\|X\|_f^2 = \frac{1}{d}\Tr(X^* X)$ denote the normalized Hilbert-Schmidt norm.

\begin{definition}[Almost homomorphism]\label{def:approx-hom-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group and $\mu_R$ a distribution on $R$. An $(\eps,\mu_R)$-almost homomorphism of $G$ is a homomorphism $\phi:\mF(S)\to\mU(\C^d)$ for some $d\geq 1$ such that
\[ \Es{r\sim \mu_R} \big\|  \phi(r) - \Id \big \|_f^2 \,\leq\, \eps\;.\]
{\color{red} YOU ALREADY SAY THAT YOU ARE GIVEN A HOMOMOTPHISM FROM THE FREE GROUP. SO $\phi(r)$ IS WELL DEFINED... NO NEED FOR THE REMARK "Here ...." AT THE END}\tnote{Ok I erased it. Michael, can you mark your comments using the command ``$\backslash$mnote$\{\cdots\}$'', {\huge AND NOT USE CAPS PLEASE}}.
\end{definition}

As already mentioned this definition makes two importance choices: firstly, to measure closeness in the Hilbert-Schmidt norm, and secondly, to measure it on average over the choice of a relation. Next we give our definition for a finitely presented group to be stable; see Definition~\ref{def:eff-stab} for the general setting.

@@

\begin{remark}
\hnote{promoted this from footnote to remark}\tnote{Can we cancel this change? It's a remark below, see Remark~\ref{rk-mus}. But in the intro, it doesn't seem so crucial to highlight} It is natural, and will always be the case for us, to take $\mu_S$ as the ``marginal'' of $\mu_R$, i.e.\ the distribution on $S$ obtained by sampling $r\sim\mu_R$ and returning a uniformly random $s$ that appears in $r$, when $r$ is considered as a multi-set.
\end{remark}

@@

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? One can verify that with $\mu_R$ and $\mu_S$ the uniform distribution on $R$ and $S$ respectively, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for some $\delta=O(\eps)$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$ \hnote{what's $\rho$? it should be related to $\delta$ somehow..}\tnote{It's whatever the soundness is. Implicitly, $\rho$ is a positive constant independent of $n$, which shows up in the constant implicit in the $O(\cdot)$ of the $\delta=O(\eps)$ You can edit if it's not clear?}. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. However, the answer is not trivial. In particular, and perhaps surprisingly, the group $G(h)$ is not necessarily Abelian, nor finite: see Remark~\ref{rk:non-abelian}. \hnote{this seems like an indirect way to motivate why it's nontrivial... why does non-Abelian-ness mean it's difficult? is there a better way of justifying the non-triviality of this question?}

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question \hnote{I'm not so convinced we've properly justified its difficulty... }\tnote{The point is that given a presentation, identifying the group is not easy; e.g. the word problem is undecidable and so we can't straightforwardly simplify the relations in a canonical form or whatnot. We could be lucky and be in a very special situation. The non-Abelian example justifies that at least some mildly non-trivial groups show up. It's true that we didn't give that much thought to the problem, so feel free to tone down the language if you prefer?}, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the most part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$ (which are the only groups we consider in this paper) the multiplication table presentation \hnote{is this standard terminology?}\tnote{Michael says so, and it sounds pretty clear to me}, which has $|G|$ generators and $|G|^2$ relations, is $\delta(\eps)=C\eps$-stable \hnote{hmm... technically the notion of stability in GH is not the same as the one defined above, right?}\tnote{Isn't it? The only difference I see is their focus on keeping control of the dimension $d'$, which Michael calls ``strongly flexibly stable'' or something like this. We could add a comment about this later, but here we're pretty high level.} for some constant $C$ that is independent of the group. This holds even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability alluded to above.

\hnote{edited this paragraph a bit; the mention of $\Z_2^k$ seemed to pop out of nowhere, but I don't think this discussion about efficiency relies on it being $\Z_2^k$.} However the results of~\cite{gowers2017inverse,de2019operator} do not imply the same quantitative stability bounds for $G(h)$ with respect to its defining presentation~\eqref{eq:gh-intro} (even if we assume that $G(h)$ is finite \hnote{could it always be finite?}\tnote{Good question---no. It's easy to get something like $F_2$ (with order-$2$ generators) by having checks that don't cross across two different subsets of variables; then there is no commutation This is also characterized in~\cite{paddock2022arkhipov} and I added a mention of it to Remark~\ref{rk:non-abelian}}). The reason to prefer the presentation~\eqref{eq:gh-intro} as opposed to the multiplication table presentation of $G(h)$ is that~\eqref{eq:gh-intro} is much more succinct: if $n,k$ are linearly related then it has $O(k)$ generators and relations, as opposed to $2^k$ and $2^{2k}$ respectively. Such a gain is essential when one recalls the interpretation of the presentation $G(h)$ as a ``tester'' for $G$ --- the size of the presentation is then directly related to the amount of \emph{randomness} required by the tester (to sample a random relation); and in computer science applications randomness is seen as an essential resource (we discuss this more below, in the context of quantum computing).

@@

\begin{theorem*}[Main, informal]
For every integer $k\geq 1$ there is a presentation $\Z_2^k = \langle S_k:R_k\rangle$ such that $|S_k|,|R_k| = 2^{\poly\log(k)}$ and furthermore this presentation is $\delta(\eps)=\poly(\log k,\eps)$-stable. \hnote{are we assuming some default setting for $\mu_S,\mu_R,d$, etc?}\tnote{The theorem statement is existential. So, implicitly, ``there exists $\mu_S,\mu_R$''. It's not clear enough?}
\end{theorem*}

@@

\begin{remark}\label{rk:non-abelian}
There exists matrices $h$ such that $G(h)$ is not $\Z_2^k$, and in fact is not Abelian. For an example, see~\cite[Example 2.16]{paddock2022arkhipov}. For that example, $n=12$, $k=7$, and the parity check matrix $h$ can be described explicitly as follows: the $7$ rows of $h$ are indexed by the vertices of the complete bipartite graph $K_{3,4}$, the $12$ columns are indexed by the $12$ edges of $K_{3,4}$, and the entry $(i,j)$ of $h$ is $1$ if and only if the edge $j$ is incident on vertex $i$. As shown in~\cite{paddock2022arkhipov}, $G(h)$ is not Abelian. By considering $K_{3,6}$ instead of $K_{3,4}$, one in addition obtains a non-Abelian \emph{infinite} group. 
\end{remark}
